\documentclass[a4paper]{article}
% \usepackage[a4paper]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{hyperref,footmisc,eurosym}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{rotating}
\usepackage{natbib,amsmath}
\usepackage{booktabs,graphicx,pgfplots,float}
\usepackage{pdflscape}
\usepackage{nameref,tabularx}
\usepackage[normalem]{ulem}
% \usepackage{lscape}
%\bibliographystyle{dinat}
%\bibliographystyle{chicago}

\linespread{1.2}
\frenchspacing
\setlength{\topmargin}{0in}
\setlength{\headheight}{0.15in}
\setlength{\headsep}{0.15in}
\setlength{\textheight}{9in}
\setlength{\textwidth}{5.5in}
\setlength{\oddsidemargin}{0.35in}
\setlength{\evensidemargin}{0in}
\setlength{\parindent}{0.15in}
\setlength{\parskip}{0.1in}
%\setlength{\voffset}{0.2in}

%%% for flow chart; start
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{caption}
\newcommand*{\h}{\hspace{5pt}}% for indentation
\newcommand*{\hh}{\h\h}% double indentation
%%% for flow chart; end


\title{Clio Infra Website:\\ Creation, Use and Maintenance.}
\author{Michail Moatsos\footnote{Researcher at the Economic and Social History Group, Department of History and Art History, Utrecht University, e-mail: m.moatsos@uu.nl}}

\begin{document}

\maketitle 
\begin{abstract}
This document contains everything one needs to know about how the new Clio Infra website is created and how to recreate it from scratch if necessary. All the requirements are detailed, while all materials and software used are open source. The main part of the procedure is build in R-scripts. Those scripts are used to fill-in manually constructed html templates. The result is a set of static pages for the web server. The R code is fed with both data and metadata that are available at the \href{https://datasets.socialhistory.org/dataverse/clioinfra}{Clio Infra dataverse repository}. Some additional metadata are drawn from other online open sources, which are detailed in the document. The html layout is made using \href{http://getbootstrap.com/}{Bootstrap}. The demo front end is reachable \href{clio2.sandbox.socialhistoryservices.org}{here}, and the official \href{https://www.clio-infra.eu}{here}. Some pending development and debugging tickets for the website and the service improvement are available on \href{https://github.com/CLARIAH/wp4-requirements/issues?q=is:issue+is:open+label:clio-infra:website}{Github}. And the Github directory for the elements necessary to rebuild the website from scratch are available \href{https://github.com/CLARIAH/wp4-clioinfra}{here}.
\end{abstract}

\clearpage

\tableofcontents

\clearpage

\section{Introduction}\label{sec:intro}

Various considerations made it necessary to re-create a more focused version of 
the Clio Infra website, and the most resource efficient way appeared to be a 
rather simple R script that would create the html pages using the Clio Infra 
data available at its dedicated 
\href{https://datasets.socialhistory.org/dataverse/clioinfra}{Dataverse 
repository}. This approach is described in detail here so any Clio Infra 
responsible, with entry-level undedrstanding of R, to be able to update the 
content when necessary.

\subsection{Built using Bootstrap}

The website is developed using Bootstrap version 3.3.7. However, some of the 
css files have been modified for additional functionalities (such as the 
back-to-the-top button). Thus, one needs to use the Bootstrap files as they are 
found on the server or at the Github repository for smooth functioning. For the 
exact html templates developed and used see section \ref{sec:req}.

\subsection{Workflow}

\begin{center}
\captionof{figure}{Workflow for an R generated website}\label{fig:diagram}
% setting the typeface to sans serif and the font size to small
% the scope local to the environment
\sffamily
\footnotesize
% Define block styles
\tikzstyle{decision} = [diamond, draw, fill=blue!30, 
    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!30, 
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!40, node distance=3cm,
  minimum height=2em]
  
  \begin{tikzpicture}[node distance = 2cm, auto]
  % Place nodes

  \node [block] (DV) {\sout{Dataverse}};
  \node [block, below of=DV] (Data) {Data downloading};
  \node [block, below of=Data] (IList) {Indicator List};
  \node [block, below of=IList] (Rankings) {Rankings \& Performance per 
Indicator \& Country};
  \node [block, left of=Data] (LEntities) {Extraction of landed entities};
  \node [block, below of=LEntities] (OECD_UN) {OECD/UN groupings};
  \node [block, below of=OECD_UN] (CList) {Country List};
  \node [block, left of=CList] (ReGlob) {Regional and Global aggregates};
  \node [block, below of=ReGlob] (Home) {Homepage, Global, Regional pages};
  \node [block, below of=CList] (CPage) {Country pages};
  \node [block, right of=Data] (meta) {Metadata downloading (docx)};
  \node [block, right of=Rankings] (IPage) {Indicator Page};
  \node [block, right of=meta] (R) {R};
  \node [block, right of=IPage] (OtherDynamic) {News, Publications, 
Contributors};
  \node [block, below of=OtherDynamic] (OtherStatic) {Partners, About, Contact};
  \node [cloud, below of=Home] (Clio) {Github};
   \node [cloud, right of=Clio] (webserver) {webserver};
  
  % Draw edges
  
  \path [line] (DV) -- (Data);
  \path [line] (DV) -- (meta);
  \path [line] (Data) -- (IList);
  \path [line] (Data) -- (LEntities);
  \path [line] (LEntities) -- (OECD_UN);
  \path [line] (OECD_UN) -- (ReGlob);
  \path [line] (ReGlob) -- (Home);
  \path [line] (Home) -- (Clio);
  \path [line] (CPage) -- (Clio);
  \path [line] (OECD_UN) -- (CList);
  \path [line] (Rankings) -- (CList);
  \path [line] (meta) -- (IPage);
  \path [line] (IPage) -- (Clio);
  \path [line] (CList) -- (CPage);
  \path [line] (R) -- (OtherDynamic);
  \path [line] (OtherDynamic) -- (Clio);
  \path [line] (OtherStatic) -- (Clio);
  \path [line] (IList) -- (Rankings);
  \path [line] (Clio) -- (webserver);
  
  \end{tikzpicture}
\end{center}

Figure \ref{fig:diagram} shows the workflow of website creation via the various 
R scripts and the available data. The Dataverse block is striked-through to 
underline the fact that in the current version those files need to be fetched 
manually in an appropriate local folder, in order for the R scripts to access 
and process them.

\subsection{Overview of items generated by the scripts}

\begin{enumerate}
 \item Indicators per country in separate xlsx with long format, created by 
running CountryHTML\_JSON.R, having filename structure:\\
CountryName\_IndicatorName\_TerritorialRef\_BorderStart\_2012\_CCode\_XXX.xlsx\\
for example: 
Zambia\_LabourersRealWage\_TerritorialRef\_1964\_2012\_CCode\_894.xlsx\\
Exported at folder: /IndicatorsPerCountry.

 \item All indicators for a given country in one xlsx with wide and long 
format, created by running CountryHTML\_JSON.R, having filename structure:\\
CountryName\_AllIndicatorsAvailable\_TerritorialRef\_BorderStart\_2012\_CCode
\_XXX.xlsx\\ for example:
BosniaandHerzegovina\_AllIndicatorsAvailable\_TerritorialRef\_1992\_2012\_CCode
\_70.xlsx \\
Exported at folder: /CountryData.

 \item (a) A ``broad'' xlsx per indicator using all ``2012 border'' countries 
of the Clio Infra dataset, and (b) a ``compact'' xlsx per indicator with rows 
only for countries with available data. Both are created by using 
IndicatorHTML.R, with both wide and long formats.\\
Exported at folder: /data.

 \item One html page for each indicator\\
Exported at folder: /IndicatorPagesWithMenus.

 \item One html page for each country\\
Exported at folder: /CountryPagesWithMenus.

 \item index.html, indexOECD.html, Partners.html, News\_Publications.html 
located at the same folder as the R scripts.
  
 \item The html files that actually need to be uploaded on the server are 
located in the folder ``ToUpload''.

\item An xlsx file containing all indicators for historical entities at 
``DataAtHistoricalBorders.xlsx''.\\
Exported at folder: /data.
 
\end{enumerate}

\textbf{WARNING: if at any point the xlsx files fail to export in the r scripts 
replace write.xls with write.xls2. Do note that most of them have been already 
replaced.}

\clearpage

\section{Requirements in tools and data}\label{sec:req}

The website is created by using a set of templates and some scripts written in 
R. The templates are written in html and make use of the css functionalities of 
\href{http://getbootstrap.com/}{Bootstrap}. 

Those HTML templates are:
\begin{itemize}
 \item IndexTemplate.html
 \item IndicatorsTemplate.html
 \item CountryTemplateJSON.html and CountryTemplateNoMoreVisualsJSON.html
 \item NewsPublicationsTemplate.html
 \item PartnersTemplate.html
\end{itemize}

The R scripts are (in strict order of execution):
\begin{itemize}
 \item ReadData.R
 \item IndicatorHTML.R
 \item CountryHTML\_JSON.R
 \item HistoricalBordersData.R
 \item AddMenusToIndicatorAndHome.R
 \item AddCountryMenu.R (only run this after running 
AddMenusToIndicatorAndHome.R)
 \item Contributors.R this needs to run only once to create 
ContributorsList.xlsx; so if there is no new contributor there is also no need 
to run this script again
 \item NewsAndPublications.R
 \item PartnersPage.R
 \item FooterSubstitution.R
 \item some supportive scripts that only need to be called from the other 
scripts are (this obviously means that \textit{you don't need to call them 
directly}): IndicatorHTMLSubstitutions.R, LongToClio.R, citations.R, 
MakeTheGGplot.R, f\_IndicatorMenu.R, and f\_IndicatorMenu2.R.
\end{itemize}

Some main texts and external images used in the process are:
\begin{itemize}
 \item AboutClioInfra.txt
 \item logo of various participating entities (found in images folder under www)
\end{itemize}

To run the aforementioned R scripts you need to install the following R 
libraries: \textit{readxl, xlsx, ggplot2, jsonlite, tidyr, RefManageR, plyr, WriteXLS, stringr, 
foreign, countrycode}. The scripts execute successfully with R version 3.6.3 (2020-02-29).

Files related to Bootstrap that need to be copied to the server www folder are 
all placed in the www folder of the Github repository. This part is excecuted 
automatically by the updating script that fetches the elements from Github and 
places them in the production server at IISH.

A few additional data files are required, which are not included in the 
dataverse repository, but they are included in the Github repository:
\begin{itemize}
 \item ``CIA-Factbook-Countries with notes for their independence status.xls'', 
constructed with information found in: 
\url{
https://www.cia.gov/library/publications/the-world-factbook/fields/2088.html#af}
 \item ``statcan - countries list and codes from statistics canada.xls'', 
constructed with information found in: 
\url{http://www.statcan.gc.ca/eng/subjects/standard/sccai/2011/scountry-desc}
 \item ``UN Countries or areas, codes and abbreviations.xls'', constructed with 
information found in:
\url{http://unstats.un.org/unsd/methods/m49/m49alpha.htm}
\item ``oecdregions.csv'' file containing the OECD split of the world intro 
regions. Provided by \href{https://www.uu.nl/staff/ARijpma}{Auke Rijpma}.
\end{itemize}

\section{Processing}

As already mentioned, the data (and metadata) files need to be manually placed 
in the appropriate folders so that R scripts can locate them. \textbf{MMM 
mention those folders explicitly here} However, in future versions of the 
scripts the data files will be fetched automatically from the dataverse.

\subsection{Metadata files}

For ease of processing and until (and if) the metadata files on dataverse are updated 
with new ones, the pre-edited metadata files found on this Github repository 
must be fed to the R scripts instead.

Because the R script requires the metadata in txt format \textbf{follow the instructions} found in ConvertDOCStoTXT.xls to convert the docx files to txt. For convenience the txt files are also provided in the Github repository.

NOTE: the names of the contributors in the metadata file need to be seperated with ``and'' or a comma.

\subsection{The website layouts}

All the HTML layouts (listed in section \ref{sec:intro}) need to be placed in 
the same folder where the R scripts are located.

\subsection{Exporting data}
 The data exporting takes place in 
\begin{verbatim}CountryHTML_JSON.R\end{verbatim} and can 
be found right below the comment within that script: 
 \begin{verbatim}
  ### Now exporting indicator and country files
 \end{verbatim}

Change the value of \textcolor{red}{ExportData} flag to export the files or 
not. If no change takes place in the data, then to considerably save time set 
this variable to \textcolor{blue}{FALSE}.

\section{Other pages}
\subsection{News \& Publications page}
This page is manually. When new items are added then they have to be 
written in HTML code within the News\_Publications.html page.

\subsection{Partners page}

Some preprocessing to extract the list of contributors is necessary. This is 
done by Contributors.R script. This needs to run only once to create 
ContributorsList.xlsx. After that only calling the PartnersPage.R is enough to 
create the Partners HTML page. Then of course you need to run the 
FooterSubstitution.R script to properly set the footer, as indicated in the 
sequence of script execution above.

\section{Adding new countries and indicators}

This section explains how to update the files and scripts when one needs to add 
one or more indicators or one or more countries.

\textbf{First step} is to place the new data and metadata files in the proper folders (doc, txt and xlsx should be in the folder historical.all.standardized.2). Remember: the first cell of the xls should have the full name of the indicator.

In case of a brand new indicator (i.e. not an update of a previously available indicator) also convert the metadata doc file that the creator of the dataset has filled to a txt file. To do so see the instructions listed in the ConvertDOCStoTXT.xls (or use simple copy paste to a txt file). Those commands are intended for use in Ubuntu. If you are not using such a system, then you will have to do this manually using for example OpenOffice under Windows. Also make sure that the data are given in the proper xls(x) layout (see file ``ClioLayout.xlsx''). If you have received them in a non-Clio layout, then you have to convert them manually. There is a script to convert the ``How was life? 2.0'' data to the Clio Layout that can be expanded accordingly (``ConvertToClioInfraLayout.R''). After the conversion you need to add two lines on top and in the first cell of the first line place the name of the indicator.

If the metadata are not given in the format required (see current txt files for 
the format required) then the process will not conclude and an error will 
occur. Thus, some manual editing of the metadata file to achieve conformity 
might be necessary. Currently the metadata template I'm using is slightly 
different than the one used previously. Thus I had to downgrade the latest 
template to achieve conformity with the old one. This means that at some point 
I need to update the procedure that will allow for the new template to be used.

 Before proceeding to the next step some modifications are necessary. Go to section starting with the 
comment: ``ADD WebName and WebCategory'' of the \textit{ReadData.R} script and add those two variables 
(ClioMetaData\$WebName and ClioMetaData\$WebCategory) for the new dataset you 
want to include (the first being the name shown on the website for the new 
dataset, and the second representing the category that this new dataset belongs 
to). Example:

\begin{verbatim}
	ClioMetaData$WebName[which(ClioMetaData$title=="Composite Measure of 
	Wellbeing")] <- "Composite Measure of Wellbeing"
	ClioMetaData$WebCategory[which(ClioMetaData$title=="Composite Measure of 
	Wellbeing")] <- "Demography"
\end{verbatim}

For ease of use keep the same WebName as it is given in the original xlsx top cell, or change that named to a one that is common in among those variables. If you do not do so, naming consistency problems arise among the various tables in the scripts.

\textbf{Second step} is that the \textit{ReadData.R} script needs to be run to read the new data in.

Before going to the third step you need to upload the new indicator on dataverse to be able to get the citation info. After depositing the new indicator on dataverse you need to download the three types of available citations to the local citation folder\footnote{.../Clio Infra/Website/Citations}. Be careful to follow the same filename structure as the other files in the Citations folder. Also update the entries in the ``~/Clio Infra/Website/CitationsStatic.xls'' file. %Then to add the bib type of citation you need to either do that manually, or install ``sudo apt install bibutils'' on an Ubuntu system. To get the command for doing this simply paste the complete short description from dataverse to the third column in ``~/Clio Infra/Website/Citations.xls'', and then export a static version of this in ``~/Clio Infra/Website/CitationsStatic.xls''.\footnote{Note that when working with the citation files from the Composite Wellbeing Index the bibutils command did not work, so I produced the bib file manually so that I can proceed.} 

Moreover, you need to add the document filename of the new indicator in the ``~/Clio Infra/Website/IndicatorsListWithDocFiles.xlsx'', and appropriately in ``~/Clio Infra/Website/IndicatorsGraphType.xlsx''; and in ``~/PhD/Clio Infra/Website/IndicPriorityList.xlsx'' set the priority in selecting the variable for country visualizations and the indicator name ``translation'', short name, short menu name and WebName.

\textbf{Third step} requires running the IndicatorHTML.R script. 

Important for completing successfully the procedure: the consistent naming of 
the indicator in various locations is critical. The filename of the xlsx with 
the original data must end with ``-historical.xslx''. The prefix (e.g. 
CompositeWellbeingIndex) should be used as it is in the docx and the txt files. 
It is preferable that the prefix will also be used in the ``ADD WebName and 
WebCategory'' above. Very important: the WebName must be used in the 
Citations.xlsx list above.

Then follow and execute the remaining scipts in order of appearance in the R script list of section \ref{sec:req}

Running ReadData.R will execute in about 2 minutes on a good laptop, and 
IndicatorHTML.R will do so in approximately 10 minutes (with all flags marked 
as True). CountryHTML\_JSON.R takes about 45 minutes with all flags up.

In the section that begins with ``\# PARAMETERS:'' in script CountryHTML\_JSON.R
you can set the parameters with which the data for the visualizations in the 
country HTML pages will be selected.

Note that when adding a historical entity a different process needs to be 
followed in script: AddMenusToIndicatorAndHome.R

To avoid reproducing the historical data files set CreateHistoricalDataFiles to 
false in script: AddMenusToIndicatorAndHome.R

\textbf{Manually change citations for Historical Gender Inequality Index}: this means
until the script is updated to do this automatically, in every total export of the data, the citation needs to be manually changed again to: ``Dilli, Selin, Sarah G. Carmichael, and Auke Rijpma. 2019. `Introducing the Historical Gender Equality Index'. Feminist Economics 25 (1): 31 - 57. https://doi.org/10.1080/13545701.2018.1442582''

\section{Uploading to the server}

\subsection{Folder correspondence table}

\begin{table}[h]
\centering
\caption{Where on server to upload which items from which local folder.}
\label{tab:uncertainty}
\begin{tabular}{l | l}
  \toprule
  Local Folder & Server Folder (under /var/www) \\
  \midrule
  ToUpload (no subdirectories) & / \\
  ToUpload/CountryPagesWithMenus & /Countries \\
  ToUpload/IndicatorPagesWithMenus & /Indicators \\
  CountryData & /docs \\
  JSON & /json \\
  IndicatorsPerCountry & /IndicatorsPerCountry \\
  data & /data \\
  html/graphs & /graphs \\
  Citations & /Citations \\
  \bottomrule
\end{tabular}
\end{table}

See table \ref{tab:uncertainty}. \textbf{Do note that the docx and xlsx files with the original data as 
deposited 
by the contributors must also be placed in the /docs folder of the server}

\iffalse
\subsection{Commands to upload to initial test server}
\begin{verbatim}
ssh michalism@clio2.sandbox.socialhistoryservices.org

scp -r "/var/www/html/theme.css" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp theme.css /var/www/

scp -r "/home/michalis/PhD/Clio Infra/Website/CountryPagesWithMenus" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp CountryPagesWithMenus/* /var/www/Countries/
scp -r "/home/michalis/PhD/Clio Infra/Website/JSON" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp JSON/* /var/www/json/
scp -r "/home/michalis/PhD/Clio Infra/Website/IndicatorsPerCountry" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R IndicatorsPerCountry/ /var/www/

scp "/home/michalis/PhD/Clio Infra/Website/index.html" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp index.html /var/www/
scp -r "/home/michalis/PhD/Clio Infra/Website/html/graphs" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R graphs/ /var/www/
scp -r "/home/michalis/PhD/Clio Infra/Website/IndicatorPagesWithMenus" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R IndicatorPagesWithMenus/* /var/www/Indicators/
scp -r "/home/michalis/PhD/Clio Infra/Website/Citations" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R Citations/* /var/www/citations/
scp -r "/home/michalis/PhD/Clio Infra/Website/data" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R data/ /var/www/
scp -r "/home/michalis/PhD/Clio Infra/Website/CountryData" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R CountryData/* /var/www/docs/

scp "/var/www/html/index.html" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
scp -r "/var/www/html/images" 
michalism@clio2.sandbox.socialhistoryservices.org:/home/michalism
sudo cp -R images/* /var/www/images/
\end{verbatim}
\fi

%\section{Bibliography}

%\bibliographystyle{apalike}
%\bibliography{library.bib}

\section{Appendix}
\subsection{Functions and Scripts}
\subsubsection{LongToClio.R}
This function converts long format to clio infra format
\subsection{MadToClio.R}
This function converts the format used in the Maddison project, to that
used in the clio-infra website and dataverse. This function was written when 
adding the Composite Measure of Wellbeing, and it is written around it, thus it 
is not a generic script yet.
\subsubsection{Contributors.R}
Exports the list of authors with their homepage and affiliation to 
ContributorsList.xlsx. The correct path needs to be provided.
\end{document}
